---
title: "Data cleaning, Tabular Formation, Chi-Squared Testing"
author: "Yoonhee Cha"
date: "`17/04/2024`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(readxl)
library(janitor)
library(skimr)
library(naniar)
library(visdat)
library(knitr)
library(kableExtra)
library(rmarkdown)
```

## Load data

```{r}
#control_df <- read_csv("control_all.csv")%>%
#                           clean_names()

pan_df <- read_excel("tooth_data.xlsx")%>%
                            clean_names()
```
Plan:

1. Get the data to tidy - every column is a variable
2. Then join on the study_id column


## Review (high level)

```{r, eval = FALSE}
# this took SO long I turned it off
#skim(control_df)
```

```{r, eval = FALSE}
#skim(pan_df)
```


Unique study participants and their groups:

```{r}

## Note I checked that number of unique IDs is equal to number of rows. When this is not the case(as below) this is a bit trickier

n <- length(unique(pan_df$study_id))


tab_cc <- pan_df %>%
          group_by(case_control)%>%
            summarise(n = n())

tab_gender <- pan_df %>%
                  group_by(case_control, gender)%>%
                      summarise(n= n())
```

There are `r n` unique study IDs in the `pan_df` dataset.


```{r}
tab_cc %>%
  kable(caption = 
          "Case control breakdown")%>%
  kable_styling()
```


```{r}
tab_gender %>%
 
  kable(caption = 
          "Case control and gender")%>%
  kable_styling()
```

## Make tidy.

This data is messy AF.

Alot of the columns can be moved from wide - long as they are just different outcomes of one variable.

Starting with `pan_df` becauase it's smaller.

The trick is `pivot_longer()`. I'm going to spell this out for you because it's handy!


First - let's find out what we're in for here
```{r}

# Any columns that are completely missing?

vis_dat(pan_df)
```

There's a whole bunch of columns towards the end that are missing. I'm going to subset the dataframe to the last 25% of columns(or whatever is convenient) so we can get a closer look.

```{r}

pan_df[,35:ncol(pan_df)]%>%
          vis_dat()

```

OK so everything from `..60` to `congenitally_missing_48` is completely empty. We'll get rid of those - no variation to speak of.

Looks like `high_bp` is missing a lot too.

Let's start with thyroid variables. We can get this into one variable - column name thyroid with values being the thyroid_ variables and a column name `thyroid_issue_present` for the row values from the original one.

The 'algorithm' is to pick columns that could be grouped together as one variable and make them into just two - column names and their values.

Where I find that the text is repetitive I go in and do a bit of tidying up there.


```{r} 
pan_long <- pan_df %>%
                select(-(x60:congenitally_missing_48))%>%
                pivot_longer(cols = thyroid_condition:unknown_thyroid_type, # columns to be switched from wide to long
                             names_to = "thyroid_condition", # what we call the column containing the names of the original column
                             values_to = "thyroid_condition_present")%>% # column we will put the values in.
                pivot_longer(teeth_anomalies:transposed_teeth, #here is where something went wrong --> needs to be anomalies but in tooth_anomalies, the tooth_damage data is populating --> why?
                             names_to = "tooth_anomalies_type",
                             values_to = "tooth_condition_present")%>% # let's clean these up a little so it's easier to understand but different from tooth
                mutate(tooth_anomalies_type = str_remove_all(tooth_anomalies_type, "teeth_"),
                       tooth_anomalies_type = str_remove_all(tooth_anomalies_type, "_teeth"),
                       tooth_anomalies_type = str_replace_all(tooth_anomalies_type, "_", " "))%>%
                pivot_longer(caries:tooth_grinding, #Tooth damage overall frequency T to X
                             names_to = "tooth_damage_type",
                             values_to = "tooth_damage_present")%>%
                mutate(tooth_damage_type = str_remove_all(tooth_damage_type, "tooth"),
                       tooth_damage_type = str_replace_all(tooth_damage_type, "_", " "))%>%
                pivot_longer(any_perio_bone_loss:any_bone_pathology,
                             names_to = "bone_loss_condition",
                             values_to = "bone_condition_present")%>%
                mutate(bone_loss_condition = str_remove_all(bone_loss_condition, "bone_loss"),
                       bone_loss_condition = str_replace_all(bone_loss_condition, "_", " "),
                       bone_loss_condition = str_replace_all(bone_loss_condition, "-", " "),
                       bone_loss_condition = str_remove_all(bone_loss_condition, "general_"),)%>%
                pivot_longer(l_carotid_calcification:stylo_hyoid_calcification,
                             names_to = "calcification",
                             values_to = "calcification_present")%>%
                mutate(calcification = str_remove_all(calcification, "_calcification"),
                       calcification = str_replace_all(calcification, "_", " "))%>%
                pivot_longer(l_throid_calc:r_thyroid_calc,
                             names_to = "thyroid_calc_location",
                             values_to = "thyroid_calc")%>%
                mutate(thyroid_calc_location = str_remove_all(thyroid_calc_location, "_thyroid_calc"))%>%
                # pivot_longer(c(osteoporosis_risk:osteo_risk_level, osteoporosis_or_osteopaenia),
                #              names_to = "osteoporosis_condition",
                #              values_to = "osteoporosis_present") %>%
                #  pivot_longer(any_cardiovascular:high_cholesterol,
                #               names_to = "heart_bp_condition",
                #               values_to = "heart_condition_presence")%>% #commented this section out bc not necessary for analysis and was making the function unable to run
                  distinct()
```


! Warning:! this data is now very long. So the `study_id` will be repeated many times. Bear this in mind when you are making tables etc.


```{r trying frequency tables}
# tab <- pan_long %>%
#   select(study_id, case_control, tooth_anomalies, tooth_condition_present)%>%
#   distinct()%>% ### just count the distinct IDs
#   filter(tooth_condition_present %in% c("Yes", "No")) %>%
#   group_by(tooth_anomalies, case_control, tooth_condition_present)%>%
#   summarise(n = n())%>%
#   ungroup()
# 
# # check sums to 183
# 
# tab <- tab %>%
#   group_by(tooth_anomalies) %>%
#   mutate(percentage = round(n/sum(n) * 100, 3),
#          cumulative_percentage = round(cumsum(percentage),3))
# 
# #tab %>%
# #  summarise(n = sum(n)) check passed
# 
# tab%>%
#   kable(caption = "Frequency of tooth anomalies")%>%
#   kable_styling()
```

\usepackage{caption}
\captionsetup[table]{labelformat=empty}

## DAMAGE
```{r tooth damage freq table}

tab <- pan_long %>%
  select(study_id, case_control, tooth_damage_type, tooth_damage_present)%>%
  distinct()%>% ### just count the distinct IDs
  #filter(tooth_condition_present %in% c("Yes", "No")) %>%
  group_by(tooth_damage_type, case_control, tooth_damage_present)%>%
  summarise(n = n())%>%
  ungroup()

tab <- tab %>%
  group_by(tooth_damage_type) %>%
  mutate(percentage = paste0(round(n/sum(n) * 100), "%"),
         cumulative_percentage = paste0(round(cumsum(n/sum(n)*100)), "%"))
  
#tab %>%
#  summarise(n = sum(n)) check passed

tab%>%
  kable(caption = "Frequency of tooth damage", row.names = FALSE)%>%
  kable_styling()
```

Question  - should checked / unchecked be removed? Or should it be Yes, No, unchecked?

```{r}
# DONT NEED TO CHECK FOR DAMAGE BECAUSE DAMAGE IS JUST A YES/NO
#library(MASS)
# check_df <- pan_long %>%
#                 select(tooth_condition_present, study_id)%>%
#                 distinct()%>%
#                 filter(tooth_condition_present %in% c("Checked", "Unchecked"))%>%
#                 group_by(tooth_condition_present)%>%
#                 summarise(n = n())%>%
#                 ungroup()
# 
# 
# check_df
```



```{r}

check_df <- pan_long %>%
                select(tooth_condition_present, study_id)%>%
                distinct()%>%
                filter(tooth_condition_present %in% c("Yes", "No"))%>%
                group_by(tooth_condition_present)%>%
                summarise(n = n())%>%
                ungroup()


check_df
```

It looks like every single study ID has been both checked and not checked? This will need some investigating and then probably something removed. The yes/no combination adds up to the correct study numbers. I would advise filtering out checked /unchecked as it's really not relevant anyway.


## PERIO
```{r perio freq table}
tab_bone <- pan_long %>%
  select(study_id, case_control, bone_loss_condition, bone_condition_present)%>%
  distinct()%>% ### just count the distinct IDs
  filter(bone_loss_condition == "any perio ") %>%
  group_by(bone_loss_condition, case_control, bone_condition_present)%>%
  summarise(n = n())%>%
  ungroup()

tab_bone <- tab_bone %>%
  group_by(bone_loss_condition) %>%
  mutate(percentage = paste0(round(n/sum(n) * 100), "%"),
         cumulative_percentage = paste0(round(cumsum(n/sum(n)*100)), "%"))
  
#tab %>%
#  summarise(n = sum(n)) check passed

tab_bone%>%
  kable(caption = "Frequency of periodontal conditions", row.names = FALSE)%>%
  kable_styling()
```

## ANOMALIES

```{r anomaly damage freq table}
tab_anomaly <- pan_long %>%
  select(study_id, case_control, tooth_anomalies_type, tooth_condition_present)%>%
  distinct()%>% ### just count the distinct IDs
  filter(tooth_condition_present %in% c("Yes", "No")) %>%
  group_by(tooth_anomalies_type, case_control, tooth_condition_present)%>%
  summarise(n = n())%>%
  ungroup()

tab_anomaly <- tab_anomaly %>%
  group_by(tooth_anomalies_type) %>%
  mutate(percentage = paste0(round(n/sum(n) * 100), "%"),
         cumulative_percentage = paste0(round(cumsum(n/sum(n)*100)), "%"))
  
#tab %>%
#  summarise(n = sum(n)) check passed

tab_anomaly%>%
  kable(caption = "Frequency of tooth anomalies", row.names = FALSE)%>%
  kable_styling()
```
## CHISQ TESTS
```{r chisq test}
#install.packages("gplots")
#library(gplots)

# do we need to do both right and left side? --> with cleaned data right and left are in the same column so would have to investigate further to determine how to do this...
chisq <- chisq.test(pan_long$thyroid_condition_present,pan_long$thyroid_calc)
chisq

# right <- chisq.test(pan_long$thyroid_condition,tooth_data$r_thyroid_calc) #right side gave warning that chi-sq approx may be incorrect so did fishers exact
# left$expected
# fisher.test(tooth_data$thyroid_condition,tooth_data$l_throid_calc)
# fisher.test(table(tooth_data$thyroid_condition,tooth_data$r_thyroid_calc))
```


```{r as word doc}
#render("Steph_Eva_work.Rmd", output_format = "word_document")
```